RationalGRL: A Framework for Argumentation and Goal Modeling

### 2018-05-07 validation.tex
Modified a bit the wordings

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a small user evaluation with 16 users. This evaluation concerned  mainly the use of the RationalGRL tool, but more general questions regarding the usefulness of arguments in goal modelling were also asked. We first present the details and results of the evaluation, and then provide a more high-level analysis in Section~\ref{sec:eval-an}.

\subsection{Experimental setup}
The study consists of three parts:
\begin{enumerate}
\item Explanation: We explain the participants the basics of the RationalGRL development process and the RationalGRL tool.
\item Modelling: We ask the participants to model a summarized discussion in the RationalGRL tool.
\item Survey: We ask some questions about the usability of the RationalGRL tool, and how it relates to other goal modeling languages.
\end{enumerate}

@@

Two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. The participants turned out to have a relatively high level of experience with IT, with an average rating of 6 years, ranging from 0 to 25 years. They also judged themselves to be quite experienced with early-phase requirements engineering (average of rating 3.2 out of 5), but experience with goal modeling was less (rating 1.9 out of 5, although two respondents use goal modeling languages almost weekly). Goal modeling languages the participants had experience with were GRL (2), i* (5), Tropos (1), and KAOS (1).

In task 2, users clearly made use of the extra features offered by the RationalGRL tool: they used the details pane more often than not (rating 3.2 out of 5) and on average produced 3 arguments from the short transcript. Some example RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}.

@@

Table~\ref{table:survey:table1} shows the respondents' answers, the average score, and the standard deviation for each answer. On average, the respondent found the tool relatively easy to use, but there was also a number of users that had some trouble getting started (in 3 out of the 4 questions, 12.\%5 rated ``very difficult''). We think this is not unexpected, as the cognitive burden for our framework is higher than for other goal modeling languages such as GRL and i*, simply because there are more elements in the language. Still, the majority of the users did not found the tool difficult to use.

We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of RationalGRL, users where generally positive about RationalGRL's clear UI and the fact that arguments can be combined with goal models. For instance, one user stated that RationalGRL successfully ``tries to capture the rationale behind the modeling process' and ``has a simple way to compute the status of the arguments''. Users did found it challenging to decide when to use arguments, and to determine when an arguments attacks another argument. For instance, user mentioned ``it wasn't very easy because I didn't always know how to attack the elements'', and ``I don't think its necessary to include all the arguments because some of them are a bit trivial''. Other users were worried about the cognitive overload of the tool. One user mentioned ``I still believe in the value of arguments, but there should be less confusing ways to capture them''.

When asked about improvements in the RationalGRL tool, the users mentioned additional UI functionality such as the possibility to save models, having an ``undo'' function, and flipping the arrow after adding them. Users also suggested various language-related improvements. Some users mentioned they missed the possibility to attack links, while other mentioned that not all GRL elements are supported (for instance, it is currently not possible to add actors).

When asked whether the users believe a more mature version of RationalGRL can be used to capture early-phase requirements in an actual software development project, most users responded positively. This was somewhat surprising, since our tool is still an early prototype. Users' responses include ``Yes, having the possibility to add arguments seems quite useful'', ``I think it can, but maybe try to find a way to combine it with more regular/mainstream requirements gathering such as user stories and customer journeys'', ``yes, because it creates a clear scope for the project, what the goals are'', and ``Generally I think its useful to explicitly document arguments of a discussion''. The worry that the modeling process may involve too much cognitive overhead was mentioned again though: ``I think the overhead of inputting a (detailed) discussion in a structured manner into any system makes adoption difficult'', and ``the manual input is too complex and takes too much time. An automated process of parsing the conversation log would be much more helpful''.

@@

We asked the users three questions about the extension of RationalGRL on a scale of 1 to 5:
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modelling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modelling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modelling?
\end{itemize}

@@

\subsection{Analysis}\label{sec:eval-an}
Although our empirical evaluation is relatively small-scale, and we cannot draw hard conclusions from it, we do believe that we can extract some interesting observations. We list the three most important ones:
\begin{enumerate}
\item \emph{High cognitive overhead.} A worry that came back often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overhead. Goal modeling is by itself already a cognitively high-effort activity, and the fact that we add more elements to the language does not improve this. This may mean that in future work we should focus more on the RationalGRL development process\footnote{Note that in order to keep the study simple for the users, we did not explicitly ask the respondents to follow the development process from Section~\ref{sect:methodology}}. Users like to be guided during modeling, and making arguments explicit in the modelling process was indicated as useful, but argumentation should be integrated into the \emph{process} of goal modeling more than in the goal models themselves. Critical questions seem to be a natural fit for this, as they play a key role in the RationalGRL development process, but in its current form they are only mentioned in the details pane of the tool.
\item \emph{Useful argumentation semantics.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool is very intuitive. This is a positive signal for our formal approach.
\item \emph{Lightweight tool.} Respondents were positive about the fact that it was very easy to get started with the tool. Thus it seems that having a simple web-interface for goal modeling languages is something that could be done more in the future. Most comments on the tool were the type of comments that are expected from a prototype implementation.
\end{enumerate}

>>>>

In addition to the empirical case study (Section~\ref{sect:gmas}), which forms the basis of the RationalGRL framework, we also performed a small user evaluation with 16 users. This evaluation concerned mainly with the use of the RationalGRL tool, but more general questions regarding the usefulness of arguments in goal modeling were also asked. We first present the details and results of the evaluation, and then provide a more high-level analysis in Section~\ref{sec:eval-an}.

\subsection{Experimental setup}
The study consists of three parts:
\begin{enumerate}
\item Explanation: We explain to the participants the basics of the RationalGRL development process and the RationalGRL tool.
\item Modeling: We ask the participants to model a summarized discussion in the RationalGRL tool.
\item Survey: We ask some questions about the usability of the RationalGRL tool, and how it relates to other goal modeling languages.
\end{enumerate}

@@

Two thirds of the participants had either a PhD or Master degree, 20\% a Bachelor degree, and one respondent responded with ``Other''. The participants turned out to have a relatively high level of experience with IT, with an average rating of 6 years, ranging from 0 to 25 years. They also judged themselves to be quite experienced with early-phase requirements engineering (average of rating 3.2 out of 5), but experience with goal modeling languages was less (rating 1.9 out of 5, although two respondents use goal modeling languages almost weekly). The participants of the study had experience with the following goal modeling languages: No experience (7 users), GRL (2 users), i* (5 users), Tropos (1 user), and KAOS (1 user).

In task 2, users clearly made use of the extra features offered by the RationalGRL tool: they used the details pane more often than not (rating 3.2 out of 5) and on average produced 3 arguments from the short transcript. Some of the examples of RationalGRL models the users created can be found in Appendix~\ref{sect:survey-screenshots}.

@@

Table~\ref{table:survey:table1} shows the respondents' answers, the average score, and the standard deviation for each answer. On average, the respondent found the tool relatively easy to use, but there were also a number of users that had some trouble getting started (in 3 out of the 4 questions, 12.\%5 rated ``very difficult''). We think this is not unexpected, as the cognitive burden for our framework is higher than for other goal modeling languages such as GRL and i*, simply because there are more elements in the language. Still, the majority of the users did not found the tool difficult to use.

We also asked the users three open questions related to usability. When asked about the strengths and weaknesses of RationalGRL, users where generally positive about RationalGRL's clear UI and the fact that arguments can be combined with goal models. For instance, one user stated that RationalGRL successfully ``tries to capture the rationale behind the modeling process' and ``has a simple way to compute the status of the arguments''. On the other hand, some of the users found it challenging to decide when to use arguments, and to determine when an arguments attacks another argument. For instance, a user mentioned ``it wasn't very easy because I didn't always know how to attack the elements'', and ``I don't think its necessary to include all the arguments because some of them are a bit trivial''. Other users were worried about the cognitive overload of the tool. One user mentioned ``I still believe in the value of arguments, but there should be less confusing ways to capture them''.

When asked about improvements in the RationalGRL tool, the users mentioned additional UI functionalities such as the possibility to save models, having an ``undo'' function, and flipping the arrow after adding them. Users also suggested various language-related improvements. Some users mentioned they missed the possibility to attack links, while others mentioned that not all GRL elements are supported (for instance, it is currently not possible to add actors).

When asked whether the users believe a more mature version of RationalGRL can be used to capture early-phase requirements in an actual software development project, most users responded positively. This was somewhat surprising, since our tool is still an early prototype. Users' responses include ``Yes, having the possibility to add arguments seems quite useful'', ``I think it can, but maybe try to find a way to combine it with more regular/mainstream requirements gathering such as user stories and customer journeys'', ``yes, because it creates a clear scope for the project, what the goals are'', and ``Generally I think its useful to explicitly document arguments of a discussion''. The concern that the modeling process may involve too much cognitive overhead was mentioned again though: ``I think the overhead of inputting a (detailed) discussion in a structured manner into any system makes adoption difficult'', and ``the manual input is too complex and takes too much time. An automated process of parsing the conversation log would be much more helpful''.

@@

We asked the users three questions about the extension of RationalGRL on a scale of 1 to 5:
\begin{itemize}
\item[Q5] Do you think the arguments and counterarguments of RationalGRL are a useful extension to standard goal modeling?
\item[Q6] Do you think the critical questions and answers in the details pane are a useful extension to standard goal modeling?
\item[Q7] Do you think the automatic determination of the status of arguments and elements is a useful extension to standard goal modeling?
\end{itemize}

@@

\subsection{Analysis}\label{sec:eval-an}
Although our empirical evaluation is relatively small-scale, and we cannot draw hard conclusions from it, we do believe that we can extract some interesting observations. We list the three most important ones:
\begin{enumerate}
\item \emph{High cognitive overhead.} A concern that was raised often in our empirical evaluation is that, in its current form, RationalGRL has a relatively high cognitive overhead. Goal modeling is by itself already a cognitively high-effort activity, and the fact that we add more elements to the language does not improve this. This may mean that in future work we should focus more on the RationalGRL development process\footnote{Note that in order to keep the study simple for the users, we did not explicitly ask the respondents to follow the development process from Section~\ref{sect:methodology}}. Users like to be guided during modeling phase, and making arguments explicit in the modeling process was indicated as useful, but argumentation should be integrated into the \emph{process} of goal modeling more than in the goal models themselves. Critical questions seem to be a natural fit for this, as they play a key role in the RationalGRL development process, but in its current form they are only mentioned in the details pane of the tool.
\item \emph{Useful argumentation semantics.} Participants were enthusiastic about adding arguments to a goal modeling language with a clear formal underpinning, and they believed the argumentation semantics we use in the tool is very intuitive. This is a positive signal for our formal approach.
\item \emph{Lightweight tool.} Respondents were positive about the fact that it was very easy to get started with the tool. Thus, it seems that having a simple web-interface for goal modeling languages is something that could be done more in the future. Most comments on the tool were the type of comments that are expected from a prototype implementation.
\end{enumerate}
